### 2017-02-24 sections/implementation.tex
Update on Overleaf.

%Application of Singular Value Decomposition to FEM results compression

@@

% jak ziskam matici na kterou aplikuji SVD? popsat z ceho se skladaji fem data a jak se naskladaji do matice

@@

Once the SVD decomposition is calculated for the result component there are two strategies how to proceed with the compression -- size or quality. Each strategy is assigned a control parameter that determines resulting compression ratio or approximation error.

\todo[inline]{Replace itemization with sub-sub-sections?}

\begin{itemize}
\item \textbf{compression ratio} -- If the focus is only on the size of compressed data, the rank $r$ of the approximation matrix can be calculated by the formula

@@

\item \textbf{approximation error} -- In the usual case the most important parameter to take into account is the approximation error. Algorithm is trying to minimize the compression ratio while at the same time ensuring that predefined limit of approximation error is not exceeded. To quantify the error the Mean squared error ($MSE$) metric is used. To be able to compare different data-sets having different scales the normalized version of $MSE$ is used - Normalized root-mean-square deviation ($NRMSD$). Both metrics are defined in section \ref{sec:error}.

But how can be the rank of approximation matrix effectively calculated from the desired approximation error? Here comes handy the interesting property of the singular values:

\begin{equation}
\sum_{i=1}^{m} \sum_{j=1}^{n} (M_{ij})^{2} = \sum_{i=1}^{k}{\sigma_{i}^{2}},
\label{eq:elem-sqr-sigma-sqr}
\end{equation}

where $k=min(m, n)$, i.e. the smallest of two dimensions of the matrix $M$. From the above follows that the squared sum of elements of a matrix $M$ equals to the squared sum of its singular values $\sigma_{i}$. \todo{Reference to source or further explanation needed.} Using formulas \eqref{eq:svd-expansion} and \eqref{eq:svd-approx-expansion} the equation \eqref{eq:elem-sqr-sigma-sqr} can be applied to the difference between original matrix $A$ and approximation matrix $A'$:

\begin{equation}
\sum_{i=1}^{m} \sum_{j=1}^{n} (A_{ij} - A'_{ij})^{2} = \sum_{i=r+1}^{k}{\sigma_{i}^{2}},
\end{equation}

where the term on the right-hand side is the sum of squares of those singular values of the matrix $A$ that are cut away by the compression algorithm. The equation can be rewritten using the definition of $MSE$ in \eqref{eq:mse-def}:

\begin{equation}
MSE \times m n = \sigma_{r+1}^{2} + \sigma_{r+2}^{2} + ... + \sigma_{k-1}^{2} + \sigma_{k}^2.
\end{equation}

@@

\todo[inline]{Describe how to calculate rank of approximation matrix - take only the amount of singular values, that the value is lower than ...}

\end{itemize}

\todo[inline]{jak se pocita chyba; jak se da ridit maximalni chyba;}

\subsection{Algorithm description}

\todo[inline]{algorithm steps ... popsat to, co se deje ve funkci Compress v SvdCompressionService, nepopisovat samotnou SVD, to by melo byt popsanot v sekci Mathematical background}

\begin{algorithm}
  \caption{Euclid’s algorithm}\label{euclid}
  \begin{algorithmic}[1]
    \Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
      \State $r\gets a\bmod b$
      \While{$r\not=0$}\Comment{We have the answer if r is 0}
        \State $a\gets b$
        \State $b\gets r$
        \State $r\gets a\bmod b$
      \EndWhile\label{euclidendwhile}
      \For{\texttt{<some condition>}}
        \State \texttt{<do stuff>}
      \EndFor
      \State \textbf{return} $b$\Comment{The gcd is b}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

@@

\todo[inline]{At first: algorithm complexity (kniha od JK os SVD - je tam slozitost algoritmu); table with execution times...}

\todo{pridat odvozeni}
Complexity: $mn^2$ where $m \leq n$ % http://mathoverflow.net/questions/161252/what-is-the-time-complexity-of-truncated-svd

\todo[inline]{Then main features for optimization: key time steps (time step span compression), Randomized SVD, Parallelization, Sparse matrix of details, prenasobeni U matice singularnimi cisly, trochu usetrim pamet...}

>>>>

@@

@@

\subsection{Algorithm description}
Once the SVD decomposition is calculated the compression algorithm removes a certain number of singular values and corresponding singular vectors. The remaining singular values and vectors represent the compressed data. There are two strategies that influence the way how to get the number of singular values to be preserved -- resulting size and quality. Each strategy is assigned a control parameter that determines compression ratio or approximation error.

\paragraph{Compression ratio}
If the focus is only on the size of compressed data, the rank $r$ of the approximation matrix can be calculated by the formula

@@

\paragraph{Approximation error}
In the usual case the most important parameter to take into account is the approximation error. Algorithm is trying to minimize the compression ratio while at the same time ensuring that predefined limit of approximation error is not exceeded. To quantify the error the Mean squared error ($MSE$) metric is used. To be able to compare different data-sets having different scales the normalized version of $MSE$ is used - Normalized root-mean-square deviation ($NRMSD$). Both metrics are defined in section \ref{sec:error}.

But how can be the rank of approximation matrix effectively calculated from the desired approximation error? Here comes useful the interesting property of the singular values

\begin{equation}
\sum_{i=1}^{m} \sum_{j=1}^{n} (\mtrx{A_{ij}})^{2} = \sum_{i=1}^{k}{\sigma_{i}^{2}},
\label{eq:elem-sqr-sigma-sqr}
\end{equation}

where $k=min(m, n)$, i.e. the smallest of two dimensions of the matrix $M$. The above formula states that the sum of squared elements of matrix $\mtrx{A}$ equals to the sum of squared singular values $\sigma_{i}$ of the same matrix $\mtrx{A}$. \todo{Reference to source or further explanation needed.} Using formulas \eqref{eq:svd-expansion} and \eqref{eq:svd-approx-expansion} the equation \eqref{eq:elem-sqr-sigma-sqr} can be applied to the difference between original matrix $\mtrx{A}$ and approximation matrix $A'$.

\begin{equation}
\sum_{i=1}^{m} \sum_{j=1}^{n} (\mtrx{A_{ij}} - \mtrx{A'_{ij}})^{2} = \sum_{i=r+1}^{k}{\sigma_{i}^{2}},
\end{equation}

where the term on the right-hand side is the sum of squares of those singular values of the matrix $A$ that are going to be cut away by the compression algorithm. The equation can be rewritten using the definition of $MSE$ in \eqref{eq:mse-def} to

\begin{equation}
MSE \times m n = \sigma_{r+1}^{2} + \sigma_{r+2}^{2} + ... + \sigma_{k-1}^{2} + \sigma_{k}^2
\end{equation}

and using \eqref{eq:nrmsd-def} further to

@@

Then $NRMSD$ can be used as a quality metric for the compression algorithm because normalization makes it usable for different datasets. Calculation of rank of the approximation matrix is depicted as pseudo-code in algorithm \ref{alg:rank-calculation}. Algorithm uses the following inequality to test whether the desired rank has been reached

\begin{equation}
e > \frac{\sqrt[]{\frac{\sigma_{r+1}^{2} + \sigma_{r+2}^{2} + ... + \sigma_{k-1}^{2} + \sigma_{k}^2}{m n}}}{X_{max}-X_{min}},
\end{equation}

where $e$ is NRMSD used as an error limit that can not be exceeded to achieve reasonable quality of approximation.

\begin{algorithm}
  \caption{Calculation of rank for approximation matrix from max allowed error}\label{rankAlgorithm}
  \label{alg:rank-calculation}
  \begin{algorithmic}[1]
  	\INPUT max allowed error ($e$), array with singular values ($S$), element count ($c$), max element value ($x_{max}$), min element value ($x_{min}$)
    \OUTPUT rank of resulting matrix
    \Procedure{CalculateRank}{$e, S, c, x_{max}, x_{min}$}
      \State $MSE \gets 0$
      \State $NRMSD \gets 0$
      \State $rank \gets S.length$
      \While{$NRMSD < e$}\Comment{repeat until max error is reached}
        \State $MSE \gets MSE + S[rank]/c$ \Comment{calculate MSE for current rank}
        \State $NRMSD \gets \sqrt{MSE} / (x_{max} - x_{min})$ \Comment{normalize error}
        \State $rank \gets rank - 1$ \Comment{decrement rank for next loop}
      \EndWhile
      \State \textbf{return} $rank + 1$ \Comment{Add one to not exceed max allowed error}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

@@

Time complexity of the exact SVD decomposition algorithm is $$O(m^2n),$$
% http://mathoverflow.net/questions/161252/what-is-the-time-complexity-of-truncated-svd

where $m<n$. This theoretical algorithm complexity is confirmed by two benchmarks where the dependency of execution time on varying matrix dimension is shown. The results of the benchmarks are depicted in Figure \ref{fig:ExeTime_rows} and Figure \ref{fig:ExeTime_columns}. Several observations were made from the results:

\begin{itemize}
\item The algorithm is most efficient in cases where one dimension of the input matrix is very low compared to the other. However, this is almost always the case when compressing results from FEM -- number of time steps seldom exceeds hundreds.
\item Moreover, time steps can be split into smaller ranges and the algorithm can be applied on each range separately. This will improve performance and can also increase quality of compression the key time steps on the range boundaries are carefully selected.
\item Randomized SVD algorithm has the same order of algorithmic complexity but even then can help to significantly reduce execution time.
\end{itemize}

Storage size of SVD decomposition itself can also be optimized in several ways. E.g. $\mtrx{\Sigma}$ being a diagonal matrix can be stored as single list of singular values $\sigma_{i}$ or can be even multiplied with the matrix of left singular vectors $\mtrx{U}$.

% main features for optimization: key time steps (time step span compression), Randomized SVD, Parallelization, Sparse matrix of details, prenasobeni U matice singularnimi cisly, trochu usetrim pamet, mohu pouzit vzorkovani...

### 2017-02-24 sections/math.tex
Update on Overleaf.

\subsection{Randomized SVD}

>>>>

\todo[inline]{Mozna tento odstavec uplne vyhodit, stejne to neni implementovano -- uvest jen v sekci Optimization nebo v Conclusion}

\subsection{Randomized SVD}
Exact SVD of a $m \times n$ matrix has time complexity $O(min(mn^2, m^2n))$ using the "big-O" notation.

Assume the data points are in the columns of $\mtrx{A} \in \mtrx{M_{m,n}}(\mathbb{R})$ where $m \leq n$. Note that $\mtrx{AA^T}$ is the dataset covariance matrix. Then a simple method is to randomly choose $k<m$ columns of $\mtrx{A}$ that form a matrix $\mtrx{S}$. Statistically, the SVD of $\mtrx{SS^T}$ will be close to that of $\mtrx{AA^T}$; thus it suffices to calculate the SVD of $\mtrx{S}$, the complexity of which, is only $O(k^2m)$.

\todo[inline]{Je tohle stejny randomizovany algoritmus jako pouzivam ja v RedSVD? V clanku se mluvi o Monte Carlo...}

% http://mathoverflow.net/questions/161252/what-is-the-time-complexity-of-truncated-svd
% http://sysrun.haifa.il.ibm.com/hrl/bigml/files/Holmes.pdf
